---
title: "The Uniform Random Variable"
output: html_document
---

&nbsp;

The uniform random variable is one of the most simplest random variables to learn. It may not seem like much but it does contain some nice properties.

&nbsp;

### <u>What is a Uniform Random Variable?</u>

&nbsp;

A uniform random variable is a random variable which takes on values from parameters $a$ to $b$ (inclusive). If the random variable $X$ is uniform then we denote it as $X \sim \text{Unif}(a,b)$.

(Note that we deal with the continuous uniform random variable and not the discrete case.)

The continuous probability distribution function (pdf) of a uniform random variable is:

&nbsp;

$$ \displaystyle f(x) = \dfrac{1}{b - a} $$

&nbsp;

If the parameters $a$ and $b$ are 0 and 1 respectively then X is a standard uniform random variable denoted as U. It would be denoted as $U \sim \text{Unif}(0,1)$. The continuous probability distribution of a standard uniform random variable is just $f(u) = \dfrac{1}{1 - 0} = 1$.

The image below shows the (theoretical) continuous probability distribution for the uniform random variable.


Source: https://www.probabilitycourse.com/images/chapter4/PDF-Uniform_b.png

&nbsp;

<u>The Cumulative Distribution Function (CDF)</u>

&nbsp;

The Cumulative Distribution Function or the CDF is the probability that a real-valued random variable $X$ with a given probability distribution is less than or equal to a quantity $x$. It is often denoted by $F(x) = P(X \leq x)$.

The CDF of a uniform random variable $X \sim \text{Unif}(a,b)$ is 

&nbsp;

$$F(x) = \dfrac{x - a}{b - a}$$

&nbsp;

for $x \in [a,b)$.

&nbsp;

**Proof**

&nbsp;

Start with the definition of the CDF along with the integral representation. Computing the integral will yield the result.

&nbsp;

$$\displaystyle F(x) = P(X \leq x) = \int_{a}^{x} \dfrac{1}{b-a} = \dfrac{x - a}{b - a}$$

&nbsp;

For the case of a standard uniform random variable, substituting $b = 1$ and $a = 0$ yields the CDF as $F(x) = x$.

&nbsp;

### <u>Generating/Simulating/Sampling Uniform Random Variables</u>

&nbsp;

Generating a uniform random variable from 0 to 1 (not including 1) is quite simple in Microsoft Excel. The function is `=RAND()`.

To generate or simulate a uniform random variable in the free statistical programming language R, we use the runif function as follows:

&nbsp;

$$\displaystyle \text{runif(n, min = 0, max = 1)}$$

&nbsp;

The $n$ in runif is the number of simulations in which the user enters. The minimum of 0 and the maximum of 1 can be altered to create a different uniform random variable.

&nbsp;

### <u>Examples</u>

&nbsp;

In these two examples, I am simulating/generating standard uniform random variables from 0 to 1 in the statistical program R.

&nbsp;

**Case One (Running 1000 simulation trials)**

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4}
data = runif(1000,min = 0, max =1)
hist(data)
```


&nbsp;

**Case Two (Running 1 Million Simulation Trials)**

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4}
data = runif(10^{6},min = 0, max =1)
hist(data)
```

Comparing the two images above, increasing the number of trials makes our sample probability distribution become closer to the rectangle shape as in the first image above.