---
title: "ASOT Top1000 Tracks Webscrape"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A State of Trance Top 1000 Tracks Of All Time
---

This is a short webscraping project where I obtain the top 1000 tracks of all time on A State of Trance radioshow hosted by Armin van Burren.



```{r cars, warning = FALSE, message = FALSE}
# A State of Trance Top 1000 Of All Time
# List Released on January 11, 2021

# Load libraries:
library(rvest)
library(dplyr)
library(tidyr)
library(stringr)
```

&nbsp;

## Extracting The Tracks
---

This code chunk below is reading the html page and extracting the tracks into the `songs` variable.

```{r asot1000}
# Link:
url <- "https://www.astateoftrance.com/news/armin-van-buuren-reveals-all-time-a-state-of-trance-top-1000-list/"

# ASOT Top 1000:
url_page <- read_html(url)

# Obtain div tag with class columns:

songs <- url_page %>%
         html_nodes("[class='columns']") 
```

&nbsp;

Tracks are found in the p tags of the songs. 

```{r}

# Tracks are found in the p tags of songs. More specifically in index 6 and 7.

tracks <- songs %>% html_elements('p') 
tracks <- tracks[c(6, 7)] %>% html_text2()
```

&nbsp;

```{r}

# Obtain top 50, extract from nested list
top_50 <-  str_split(tracks[1], pattern = "\n") 
top_50 <- top_50[[1]]

top_50

```

&nbsp;

```{r}

# Ranked 51-1000
other_tracks <- str_split(tracks[2], pattern = "\n")
other_tracks <- other_tracks[[1]]

# Combine all:
combined <- c(top_50, other_tracks)

# Dataframe:
df <- data.frame(Rank = seq(1, 1000), Track = combined)

head(df, 20)
```

&nbsp;

## Extension: Having Rank, Artist & Song As Three Separate Columns

The format before was just rank and track. This extension part will separate the artist and song from the hyphen. One small detail to note is that the hyphen from the A State of Trance list is a different hyphen from my keyboard. You have to copy and paste this hyphen in the `sep = ` argument in `separate()`.

&nbsp;

```{r}
# Separate by hyphen. That is the hyphen from the site (-) not my keyboard hyphen (-):
asot_top_1000_df <- df %>% 
                    separate(col = Track, into = c("Artist", "Song"), sep = '–')  

# Remove number in artist column:
# Reference: https://stackoverflow.com/questions/32767164/use-gsub-remove-all-string-before-first-white-space-in-r
asot_top_1000_df$Artist <- str_replace(asot_top_1000_df$Artist, "^[\\S]+", "") %>% trimws()

# Preview table:
head(asot_top_1000_df)

tail(asot_top_1000_df)
```

&nbsp;

### Most Frequent Artists In ASOT Top 1000
---

The `count()` function from `dplyr` can be used to obtain counts.

```{r}
# Top 20 Most Frequent Artists in ASOT Top 1000
# Armin van Buuren, Above & Beyond, GAIA, Giuesppe Ottaviani, Ørjan Nilsen, Aly & Fila, Gareth Emery
asot_top_1000_df %>% 
  count(Artist) %>% 
  arrange(desc(n)) %>%
  head(20)

# Save table as .csv file:
write.csv(asot_top_1000_df, file = 'ASOT_Top1000_Tracks.csv', row.names = FALSE)
```
