---
title: "A Look At R's stringr Package"
output: html_document
---

&nbsp;

Image Source: https://www.rstudio.com/wp-content/uploads/2014/06/RStudio-Ball.png

&nbsp;

**Warning** This post is very long.

&nbsp;

Hi there. I have been playing around with the stringr package in the statistical programming language R. This stringr package is a neat package for dealing with strings (text) and data cleaning/formatting.

In this page I showcase some experimentation work I have done using the stringr package function. I go over large portion of the functions in stringr but not all. (There are a few functions that I have not tested or not familiar with.)

My reference here is the R documentation on stringr and its functions.

Note: I am not that familiar with regular expressions but I am learning about it through this page.

&nbsp;

### <u>Sections</u>


&nbsp;

- The stringr Package In R

- Installing and Loading stringr

- A Look At The stringr Functions
    - <a href="#length">Length Of Strings</a>
    - <a href="#convert">Converting To Uppercase, Lowercase, Titles</a>
    - <a href="#combine">Combining Multiple Strings Into A Single String</a>
    - <a href="#matches">The Number Of Matches In A String</a>
    - <a href="#detect">Detecting Matches In A String</a>
    - <a href="#duplicate">Duplicating & Multiplying Strings</a>
    - <a href="#extract">Extract A Matching Pattern From A String</a>
    - <a href="#position">Locating The Position Of A Pattern In A String</a>
    - <a href="#match">String Matching</a>
    - <a href="#sorting">Ordering/Sorting Strings</a>
    - <a href="#padding">Padding Strings</a>
    - <a href="#replace">Replacing Patterns In Strings</a>
    - <a href="#split">Splitting Strings</a>
    - <a href="#substrings">Substrings From Strings</a>
    - <a href="#trimming">Trimming Strings</a>
    - <a href="#truncating">Truncating Strings</a>
    - <a href="#words">Extracting Words From A Sentence</a>

&nbsp;


### <u>The stringr Package In R</u>

&nbsp;

There are cases where you may need to manage strings/text in R. This stringr package is useful for dealing with such strings.

In R/RStudio, you can type in `??stringr` and then click on `stringr::stringr` from the help window to find out some information on stringr. Other resources on stringr include the R documentation here (.pdf) and this webpage.

&nbsp;


### <u>Installing and Loading stringr</u>

&nbsp;

To install the stringr package, type in `install.packages("stringr")` into R/Rstudio.

After installing the stringr package into R, you can load the stringr package by typing in 

&nbsp;

```{r}
library(stringr)
```

&nbsp;

Once the stringr package is loaded, the functions in the stringr package can be used.


&nbsp;

### <u>A Look At The stringr Functions</u>

&nbsp;

There are quite a few functions in the stringr package so I am keeping details at a minimum. Each subsection refers to a specific function in stringr. Examples are included to show how the functions operate.

&nbsp;

<a name="length"></a>

**Length Of Strings**

&nbsp;

The length or number of characters in a string can be determined with the str_length() function from stringr. Please refer to the examples below.

&nbsp;

```{r}
## Length of strings:

str_length("Hello")

str_length(c("Green", NA, "R", "Programming"))

```

&nbsp;

<a name="convert"></a>

**Converting To Uppercase, Lowercase, Titles**

&nbsp;

Converting strings to uppercase/lowercase letters or to a title format is not difficult. The functions involved are `str_to_upper()`, `str_to_lower()` and `str_to_title()`.

&nbsp;

```{r}
## Converting cases in a string:

word <- "Green"

print(str_to_upper(word)) # Uppercase

print(str_to_lower(word)) # Lowercase

print(str_to_title(word)) # Title, first letter upper case, rest lowercase
```

&nbsp;

<a name="combine"></a>

**Combining Multiple Strings Into A Single String**

&nbsp;

You can combine/concatenate multiple strings into a single string with the `str_c()` function. According to the R documentation, the function has this format.

&nbsp;

```
str_c(..., sep = "", collapse = NULL)
```

&nbsp;

Here are some examples.

&nbsp;

```{r}
## My example:

fibonacci_ten <- c("1", "1", "2", "3", "5", "8", "13", "21", "34", "55")

str_c("Number: ", fibonacci_ten)

str_c("Number", fibonacci_ten, sep = " - ")

str_c("The number ", fibonacci_ten, " is a Fibonacci number.")

str_c(fibonacci_ten[-10], " is less than or equal to ", fibonacci_ten[-1])

str_c(c("1", NA, "10"), "2") # missing values cases = missing outputs, concatenate strings

```

<a name="matches"></a>

&nbsp;


**The Number Of Matches In A String**

&nbsp;

The `str_count()` function can help in finding the number of a specific pattern in a string. Examples are shown below.

```{r}
## Number of matches in a string using str_count():

fruit <- "strawberry"

str_count(fruit, pattern = "r") # Find r's


str_count(fruit, pattern = "w") # Find w's


str_count(fruit, pattern = c("w", "a", "y", "r")) # Find w's, a's, y's and r's

colours <- c("green", "red", "blue", "purple", "orange", "yellow", "teal")

str_count(colours, pattern = "e") # Find number of e's in colours; output is a vector

str_count(colours, pattern = "l") # Find number of l's in colours

```

&nbsp;

<a name="detect"></a>


**Detecting Matches In A String**

&nbsp;

In stringr, you can detect certain matches/patterns in a string with the `str_detect()` function. Knowledge of regular expressions is helpful depending on what you want to detect. Outputs from `str_detect()` are either TRUE or FALSE.

&nbsp;

```{r}
## Detect matches in a string using str_count(). Output is a logical vector.
## Regular expressions can be used too.

item <- "chair"

str_detect(item, pattern = "a") # Is there an a?

str_detect(item, pattern = "y") # Is there a y?

str_detect(item, pattern = "[aeiou]") # Is there a vowel?

str_detect(item, pattern = "^c") # Is there a starting c?

str_detect(item, pattern = "c$") # Is there an ending c?

str_detect(c(item, "TV"), pattern = "^c") # Is there a starting c?
```

&nbsp;

<a name="duplicate"></a>

**Duplicating & Multiplying Strings**

&nbsp;

You can create duplicates and multiply strings together with the `str_dup()` function.

&nbsp;

```{r}
## Duplicating and combining strings:
# These examples are similar to the R documentation examples.

food <- c("Pizza", "Chicken", "Noodles", "Salad")

str_dup(food, 2) # Double

str_dup(food, 1:4) # Single for Pizza, 2x Chicken, 3x Noodles, 4x Salad

str_c(food, str_dup("!", 1:4)) # Add exclamation marks to food items.

```

&nbsp;

<a name="extract"></a>
&nbsp;

**Extract A Matching Pattern From A String**

The `str_extract()` function from the stringr package allows for extracting certain patterns from a string. You can use a single string or a vector of strings in `str_extract()`. Regular expressions can be used as well here. More examples can be found in the stringr documentation.

&nbsp;

```{r}
## Extract the matching pattern from string:
# Can use regular expressions; more examples in documentation.

colours <- c("green", "red", "blue", "purple", "orange", "yellow", "teal")

str_extract(string = colours, pattern = "e") # Extract letter e.

str_extract(string = colours, pattern = "[aeiou]") # Extract vowels.

str_extract_all(string = colours, pattern = "e") # Extract all cases of letter e.

str_extract_all(string = colours, pattern = "[aeiou]") # Extract all vowel cases.

```

&nbsp;

<a name="position"></a>


**Locating The Position Of A Pattern In A String**

&nbsp;

There are times when you would rather locate the position of the pattern in a string versus extracting that pattern. The `str_locate()` function requires a string and a pattern which would output the position of the specified pattern.

If the input is a vector of strings, the output would be a matrix where the number of rows is the number of strings and there would be two columns. The first column represents the starting position for a string and the second column represents the ending position of the string. If the pattern is a single character, the start and end position would be the same.

&nbsp;

```{r}
# Locating the position of patterns in a string:

colours <- c("green", "red", "blue", "purple", "orange", "yellow", "teal")

str_locate(colours, "e") # Find position of e.

str_locate(colours, "a") # Find position of a.

# Find cases of letter e in green, d in red, u in blue, p in purple and so on:

str_locate(colours, c("e", "d", "u", "p", "n", "y", "l"))

str_locate_all(colours, "e") # Find positions of e in each colour in colours.

str_locate_all(colours, "o") # Find positions of o in each colour in colours.

```

&nbsp;

If there is no such pattern, the output would be NA or blank.

&nbsp;

<a name="match"></a>


**String Matching**

&nbsp;

You can match certain patterns in a string using the `str_match()` function. This function is very similar to `str_extract()`.

&nbsp;

```{r}
## String matching:

str_match(string = "yellow", pattern = "ll")

str_match_all(string = "parallelogram", pattern = "l")

```

&nbsp;

<a name="sorting"></a>

**Ordering/Sorting Strings**

&nbsp;

A useful function is ordering/sorting strings. The functions used here are `str_order()` and `str_sort()`. Given a vector of strings, `str_order()` outputs the positions in ABC order (by default). The `str_order()` function outputs the elements in the vector in ABC order.

The template code from the R documentation for both functions are shown below. (I do not know what locale means here.)

&nbsp;

```
str_order(x, decreasing = FALSE, na_last = TRUE, locale = "", ...)
str_sort(x, decreasing = FALSE, na_last = TRUE, locale = "", ...)
```

&nbsp;

The first example uses five numbers.

&nbsp;

```{r}
## String Order/Sort:

# Template Code: str_order(x, decreasing = FALSE, na_last = TRUE, locale = "", ...)
#                str_sort(x, decreasing = FALSE, na_last = TRUE, locale = "", ...)
# Don't know what locale is.

five_numbers <- as.character(c(1:5))
five_numbers

# Order from higherst to lowest, decr is FALSE by default:

str_order(five_numbers, decreasing = TRUE) # Gives positions in vector.

str_sort(five_numbers , decreasing = TRUE) # Outputs each number in vector.

```

&nbsp;

Note: It is recommended to not sort numbered strings using `str_sort()` or `str_order()`. Here is an example.

&nbsp;

```{r}
# Example - Ordering numbers as strings is bad.

str_sort(as.character(c(1:10)), decreasing = TRUE)
```

&nbsp;

The output above does not have "10" as the first string. This "10" is second last as I think it looks at the first digit and this "10" is before "1" because of this extra digit.

These next examples are about sorting text alphabetically.

```{r}
# Ordering objects:

objects <- c("Box", "TV", "Computer", "Phone", "Oven")

# ABC order:

str_order(objects) # Gives positions in vector.

str_sort(objects) # Outputs each number in vector.
  
# ZYX Order:

str_order(objects, decreasing = TRUE) # Gives positions in vector.

str_sort(objects , decreasing = TRUE) # Outputs each number in vector.

```

&nbsp;

**Padding Strings**

&nbsp;

Padding strings involve adding characters to the left/right or both sides of a string. The code template from the R documentation is shown below along with some examples.

&nbsp;

```{r}
## Padding a string:

# Template: str_pad(string, width, side = c("left", "right", "both"), pad = " ")

str_pad("Jello", width = 7, side = "right", pad = "!")

str_pad("Jello", width = 7, side = "left", pad = "J")

# Add 2 Js both sides and 2 !s both sides a vector with 2 elements.

str_pad("Jello", width = 9, side = "both", pad = c("J", "!"))

# Making JJJello!!. Inside str_pad makes Jello!!, Outside one adds 2 Js to Jello!!:

str_pad(str_pad("Jello", width = 7, side = "right", pad = "!"),
         width = 9, side = "left", pad = "J")

```

&nbsp;

<a name="replace"></a>

**Replacing Patterns In Strings**

&nbsp;

You can replace patterns in strings with the `str_replace()` and `str_replace_all()` functions from stringr. The user would have to specify the pattern to be replaced inside a string (or a vector of strings) and a replacement.

&nbsp;

```{r}
## String Replace:

term <- "yellowjello"

# Replace "ll" with black space:

str_replace(term, pattern = "ll", replacement = "") # First instance

str_replace_all(term, pattern = "ll", replacement = "") # All instances

# Replace "ll" with "nn" :

str_replace(term, pattern = "ll", replacement = "nn")

str_replace_all(term, pattern = "ll", replacement = "nn")

```

&nbsp;

<a name="split"></a>

**Splitting Strings**

&nbsp;

Given a string or a vector of strings, you can split strings with the `str_split()` function. There is also the `str_split_fixed()` function where the output is a character matrix versus a character vector from `str_split()`. The code below shows the R documentation template code with some of my examples.

&nbsp;

```{r}
## String Split - Split strings into pieces.

# Template: str_split(string, pattern, n = Inf, simplify = FALSE)

str_split("Hannah Banana", pattern = " ")

str_split("ABC", pattern = "") # Split into each character

oils <- "Canola, Sunflower, Avocado, Sesame, Olive, Castor, Peanut"

str_split(oils, pattern = ", ")

str_split(oils, pattern = ", ", n = 2)

str_split(oils, pattern = ", ", n = 1) # One vector

str_split(oils, pattern = ", ", n = 3)

str_split(oils, pattern = ", ", n = 3, simplify = TRUE) # Return character matrix

str_split_fixed(string = oils, pattern = ", ", n = 3)

str_split_fixed(string = oils, pattern = ", ", n = 2)

```

&nbsp;

<a name="substrings"></a>

&nbsp;

**Substrings From Strings**

&nbsp;

Suppose you have a string (or strings) where you want a subset/portion of the string(s). The `str_sub()` function allows the R user to extract substrings from strings. Start and end positions are needed along with the string object itself in `str_sub()`. (You could also use `str_locate()` to find positions of patterns.)

&nbsp;

```{r}
## String Sub:
# Extract & replace substrings from a character vector.

name <- "Roger Federer"

str_sub(name, start = 1, end = 5) # Roger

str_sub(name, start = 7, end = 13) # Federer

# str_count(name) = 13; Output is Federer:

str_sub(name, start = 7, end = str_length(name))

# Fed:

str_locate(name, pattern = "Fed") #7 and 9:

str_sub(name, start = 7, end = 9)

```

&nbsp;

<a name="trimming"></a>


**Trimming Strings**

&nbsp;

Any whitespace in strings can be trimmed/removed by using the `str_trim()` function. The user has to specify whether to trim from the left, right or both sides of the string.

&nbsp;

```{r}
## String Trim to trim whitespace at start and end of a string:

test <- " Hi "

str_trim(test, "left")

str_trim(test, "right")

str_trim(test, "both")

test_cases <- c(" Hi ", " Hey! ")

str_trim(test_cases, "left")

str_trim(test_cases, "right")

str_trim(test_cases, "both")

```

&nbsp;

<a name="truncating"></a>

**Truncating Strings**

&nbsp;

Truncate can be seen as another word for trim or subset. In stringr, the `str_trunc()` function takes a string, a specified width (substring length), and a side of either right, left or center. The ellipsis = "..." is there by default. The output for `str_trunc()` gives a substring customized to the arguments in `str_trunc()`.

The code below includes the template code from the R documentation and some examples.

&nbsp;

```{r}
## String Truncate: Extract a number of characters from a string:

# Template: str_trunc(string, width, side = c("right", "left", "center"), ellipsis = "...")

long_sentence <- "This is a very long sentence with many different letters."

str_trunc(string = long_sentence, width = 25, side = "right")

str_trunc(string = long_sentence, width = 25, side = "left")

str_trunc(string = long_sentence, width = 25, side = "center")

```

&nbsp;

<a name="words"></a>


**Extracting Words From A Sentence**

&nbsp;

Given a string that is a sentence, the `word()` function from stringr can extract words. The user has to specify a start and end. If you want to extract the first two words in a sentence you would use start = 1 and end = 2. You can somewhat view each word in a sentence as elements in an array/vector.

&nbsp;

```{r}
## word(): Extract words from a sentence.

sentence <- "The cat in a hat is missing."

word(string = sentence, start = 1, end = 2) # Extract first two words.

word(string = sentence, -1) # Extract last word.

word(string = sentence, -3) # Extract third last word.

# Multiple sentences case:

sentence_list <- c("Watch out there!", "Here is my car.")

word(sentence_list, 1) # First word in each sentence.

word(sentence_list, start = 1:2, end = -1)

```

&nbsp;
