---
title: "R's sapply() Function With Examples"
output: html_document
---

&nbsp;

Hi there. This page is about R's sapply function. The sapply function is a neat function where values in a dataframe column can be used as inputs for functions inside sapply. This concept will make more sense with the examples below.

&nbsp;

### <u>Examples Of sapply()</u>

&nbsp;

<u>Example One</u>

&nbsp;

In the first example, I create a table of values for a simple quadratic function $y = x^2$. The x-values are natural numbers from 1 to 10 and is in a dataframe.

&nbsp;

```{r}
#### Example One:

# Create dataframe:

df <- data.frame(x = 1:10)
```

&nbsp;

A R function for the $y = x^2$ quadratic equation is made which will be used for sapply() later.

&nbsp;

```{r}
# Function for x^2:

x_squared <- function(x){
  return(x^2)
}
```

&nbsp;

Once the function is created, this function can be used in the sapply function. I apply this function to the x column such that the x-values are inputs in the x_squared function.

&nbsp;

```{r}

# Change second column from xvalues to its squared values with sapply:

df$y <- sapply(df$x, x_squared)

# View dataframe:
df
```

&nbsp;

A natural extension is to use R's `ggplot2` package to generate a plot and function sketch.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# ggplot Plot:

library(ggplot2)

ggplot(data = df, aes(x = x, y = y)) +
  geom_point() +
  geom_line(col = "red") +
  annotate("text", x = 5, y = 85, parse = TRUE, size = 7, fontface ="bold",
           label= "y == x^2") +
  theme_bw()
```

&nbsp;

<u>Example Two (Anonymous Function)</u>

&nbsp;

In the `sapply()` function, you can create a function as an argument instead of defining a function outside of `sapply()`. Refer to the example below where the function is the square root of x ($y = \sqrt(x)$).

&nbsp;

```{r}
#### Example Two (Anonymous Function):

df2 <- data.frame(x = 1:20)

# Change second column to its square root values with sapply and an anonymous function:

df2$y <- sapply(df2$x, function(x){sqrt(x)})

df2
```

&nbsp;

Like in the first example, you can use ggplot2 for plotting.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# ggplot Plot:

ggplot(data = df2, aes(x = x, y = y)) +
  geom_point() +
  geom_line(col = "red") +
  annotate("text", x = 15, y = 2, parse = TRUE, size = 7, fontface ="bold",
           label= "y == sqrt(x)") +
  theme_bw()
```

&nbsp;


<u>Example Three</u>

&nbsp;

R's `sapply()` does not necessarily have to be used for math. This example looks an numeric grades and gives a pass/fail outcome with the passFail function and `sapply()`.

(The passing grade here is 60 instead of the regular 50.)

&nbsp;

```{r}
#### Example Three:

grades <- c(62, 87, 39, 57, 80, 82, 77, 72, 70, 69, 64, 70, 59, 61)

grades_df <- data.frame(Grades = grades)

# Pass is 60 instead of regular 50.

passFail <- function(num_grade){
  if(num_grade < 60){
    return("Fail")
  } else{
    return("Pass")
  }
}

# Using sapply to determine letter grade from number grade

grades_df$status <- sapply(grades_df$Grades, passFail)

grades_df

```

&nbsp;

<u>Example Four</u>

&nbsp;

This example features assigning a weight condition given a BMI score. I have created a (fake and random) list of BMI numbers in R.

I have used this website as a reference for BMI categories.

&nbsp;

```{r}
#### Example Four (BMIs):
# Reference: https://www.nhlbi.nih.gov/health/educational/lose_wt/BMI/bmicalc.htm?source=quickfitnesssolutions

bmis <- c(17, 21.4, 18.2, 28.1, 30.6, 32.1, 22.3, 21, 23.2, 25.7, 31.7, 19.9, 26.2, 27.8)

bmis_df <- data.frame(Body_Mass_Index = bmis)
```

&nbsp;

To assign weight conditions given numeric bmis, an if-elseif-else format is needed in the function for `sapply()`.

&nbsp;

```{r}
# Assign BMI number to status condition with a function

bmi_status <- function(bmi){
  if(bmi < 18.5){
    return("Underweight")
  } else if (bmi >= 18.5 & bmi <= 24.9){
    return("Normal Weight")
  } else if (bmi >= 18.5 & bmi <= 24.9){
    return("Normal Weight")
  } else if (bmi >= 25 & bmi <= 29.9){
    return("Overweight")
  } else {
    return("Obese")
  }
}

# Using sapply to determine weight status given a BMI score in bmi column.

bmis_df$Status <- sapply(bmis_df$Body_Mass_Index, bmi_status)

# View/Check
bmis_df

```





