---
title: "Cartesian Products In R"
output: html_document
---

&nbsp;

Here is a short guide to producing Cartesian Products in R. Cartesian products is a concept from the mathematical field of set theory. To create Cartesian products at least two sets are required. Those from a computer science background would be more familiar with arrays (vectors) more than sets.

A related post to this would be one of my old pages An Introduction To Set Theory & Logic (Part Three Of Three)

&nbsp;

### <u>Table Of Contents</u>

&nbsp;

* Cartesian Products
* An Example Of A Cartesian Product
* Examples of Cartesian Products in R
* Reference

&nbsp;

### <u>Cartesian Products</u>

&nbsp;

The Cartesian product of sets $A$ and $B$ is denoted by $A \times B$ or "A cross B". It is the set of all ordered pairs (a, b) such that $a \in A$ and $b \in B$ or $A \times B = \{(a, b) : a \in A, b\in B\}$. You can also think of the Cartesian products as all possible combinations between two sets.

If we have $A \times A$ then that can be written as $A^{2}$.

If either set $A$ or set $B$ (or both of them) is empty ($\varnothing$) then the Cartesian product would be empty as well. ($A \times B = \varnothing$)

In the Cartesian product, we can determine the number of elements in this set. Denote $|A|$ as the number of elements in the set A and denote $|B|$ as the number of elements in the set B. If sets $A$ and $B$ are sets with a finite number of elements then $|A \times B| = |A| \cdot |B|$.

&nbsp;

### <u>An Example Of A Cartesian Product</u>

&nbsp;

Here is an example of how the Cartesian product operates.

Suppose we have the set $P = \{1, 2\}$ and the set $Q = \{a, b, c\}$. What is the Cartesian product $P \times Q$? What about $Q \times P$ and $P \times P$?

&nbsp;

**Answer**

$$P \times Q = \{ (1, a), (1, b), (1, c), (2, a), (2, b), (2, c)\}$$

$$Q \times P = \{ (a, 1), (a, 2), (b, 1), (b, 2), (c, 1), (c, 2) \}$$

$$P \times P = \{ (1, 1), (1, 2), (2, 1), (2, 2)\}$$

The Cartesian products here are all possible combinations between the two sets.

From the sets P and Q we can find the number of elements in the Cartesian products. The sets $P \times Q$, $Q \times P$ contain $2 \times 3 = 6$ elements each and $P \times P$ has $2 \times 2 = 4$ elements.

&nbsp;

### <u>Examples of Cartesian Products in R</u>

&nbsp;

In the statistical program R, there is a built in function which can produce Cartesian products from at least two sets/vectors. This `expand.grid()` function is the R version of Cartesian products.

The first example in R is the example from the previous section. Note that the output in R has a different order from the answers above.

&nbsp;

```{r, echo = TRUE}
# Example One:

P <- c(1, 2)
Q <- c("a", "b", "c")

# Cartesian Products (Table Form):
# First column from the first vector, second column from second vector.

prod1 <- expand.grid(P, Q)
prod1

prod2 <- expand.grid(Q, P)
prod2

prod3 <- expand.grid(P, P)
prod3

# Number of ordered pairs or elements in the Cartesian product for product4:
nrow(prod3)





```

&nbsp;

This second example is somewhat more practical in the sense that the Cartesian product helps in finding all possible combinations between the fruits set and the animals set.

&nbsp;

```{r}
# Example Two:

fruits <- c("Banana", "Orange", "Mango")
animals <- c("Dog", "Cat", "Panda", "Elephant")

product1 <- expand.grid(fruits, animals)
product1

product2 <- expand.grid(animals, fruits)
product2

product3 <- expand.grid(fruits, fruits)
product3

product4 <- expand.grid(animals, animals)
product4

# Number of ordered pairs or elements in the Cartesian product for product4:
nrow(product4)
```

&nbsp;

If you want to change the column titles from Var1 and Var2, you can use the following line of code in R:

```
colnames( table_name ) <- c("col1_title", "col2_title")
```

&nbsp;

**Reference**

Source: http://stackoverflow.com/questions/4309217/cartesian-product-data-frame-in-r
