---
title: "Plotting Uniform Distributions In R With ggplot2"
output: html_document
---

&nbsp;

Hello there. This page is about plotting uniform distributions in R with the ggplot2 package. The approach I use is through making a function where the user specifies the minimum and maximum of the uniform distribution and then the function outputs the associated uniform distribution plot.

&nbsp;

### <u>Creating The Uniform Distribution Plotting Function</u>

&nbsp;

This time around, I have decided to put ggplot2 code inside a function as the uniform distribution varies depending on the values of a and b (minimum and maximum values). The function is designed to adapt to different a and b values.

If the ggplot2 package is not installed into R/RStudio, type in `install.packages("ggplot2")` the following to install `ggplot2`.

In R, load in the ggplot2 package by typing in `library(ggplot2)`.

---

Recall that functions in R have the form of:

&nbsp;

```
function_name <- function(arg1, arg2, ..., argn){
    put code here
}
```

&nbsp;

In the code below, I have the arguments as a and b for the uniform distribution and the xvals line sets the range for the x-values. The ggplot2 lines of codes takes in xvals as the xlimits for the plot, and the uniform distribution plots and labels adjusts to the a and b values.

&nbsp;

```{r, echo = TRUE, fig.width=3, fig.height=3} 
# Plotting Uniform Distributions In R With ggplot2

# Using the ggplot package to plot various probability distributions.

# install.packages("ggplot2")

library(ggplot2)

## Uniform Distribution - Unif(a, b) or Unif(min, max):

uniform_Plot <- function(a, b){
  xvals <- data.frame(x = c(a, b)) #Range for x-values
  
  ggplot(data.frame(x = xvals), aes(x = x)) + xlim(c(a, b)) + ylim(0, 1/(b - a)) +
    stat_function(fun = dunif, args = list(min = a, max = b), geom = "area", 
                  fill = "green", alpha = 0.35) + 
    stat_function(fun = dunif, args = list(min = a, max = b)) +
    labs(x = "\n u", y = "f(u) \n", 
         title = paste0("Uniform Distribution \n With Min = ", a, " & Max = ", b, " \n")) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.title.x = element_text(face="bold", colour="blue", size = 12),
          axis.title.y = element_text(face="bold", colour="blue", size = 12)) +
    geom_vline(xintercept = a, linetype = "dashed", colour = "red") +
    geom_vline(xintercept = b, linetype = "dashed", colour = "red")
  
}
```

&nbsp;


The two `stat_function()` addons are for creating the probability density line and for the colour fill under the line. I have used $\dfrac{1}{(b - a)}$ as the upper limit for the y-limit as $\dfrac{1}{(b - a)}$ is the "height" of the uniform distribution.

In `labs()`, I have the `paste0()` function for title to adjust depending on the values of a and b. Having `paste0()` be adjustable allows for adjustable titles.

The two red geom_vlines are included to highlight the boundaries of the uniform distributions at a and b.

&nbsp;

### <u>Examples</u>

&nbsp;

Once this function is set up in R, you can make function calls with your choice of values for a and b.

```{r, echo = TRUE, fig.width=4, fig.height=4} 
## Test Cases:

# Plot uniform distribution with min = 1, max = 2

uniform_Plot(a = 1, b = 2)

# Plot uniform distribution with min = 4, max = 4.5:

uniform_Plot(a = 4, b = 4.5)

# Plot uniform distribution with min = -1, max = 1:

uniform_Plot(a = -1, b = 1)

### Standard Uniform Distribution Where it's a uniform dist. with a = 0, b = 1:

uniform_Plot(a = 0, b = 1)
```

