---
title: "Fibonacci Sequence Numbers In R"
output: html_document
---

&nbsp;

Hi. This page is about creating a function which would extract Fibonacci numbers in R.

(This function can also be coded in other programming languages such as Python and C++. )

&nbsp;

---

&nbsp;

### <u>A Function Which Extracts A Fibonacci Number In R</u>

&nbsp;

My main objective for this function is to have the function output the n-th Fibonacci number in a Fibonacci sequence.

Remember that the Fibonacci sequence starts with 1 and 1 as follows:

&nbsp;

$$1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, ...$$

&nbsp;

An alternate version of the Fibonacci sequence has the sequence starting at 0 and 1.

If $n = 5$, the fifth Fibonacci number in the above sequence would be 5.

&nbsp;
---

&nbsp;

Before introducing the Fibonacci function, I have included a brief review of functions in R, if-else statements in R and for-loops in R.

&nbsp;

### <u>Functions In R</u>

&nbsp;

Recall that a function in R follows this format.

&nbsp;

```
function_name <- function(arg1, arg2, ...) {
       <put code here>
}
```

&nbsp;

To use the function, the function_name along with the argument(s) needs to be declared. An example of a function call would be something like:

&nbsp;

```
function_name(arg1 = *, arg2 = *, ...)`
```

&nbsp;

### <u>If-Else Statements In R</u>

&nbsp;

In this Fibonacci function, I use an if-else statement for dealing with cases. I want the user of the function to put in whole numbers that are at least 1.

In R, the if-else statement follows a format which looks something like this:

&nbsp;

```
if (case_1) {
   < code here for case_1 >
} else if (case_2) {
   < code here for case_2>
} ... {

} else if (case_[n -1]){
   < code here for case_[n - 1]>
} else #case_n
   < code here for case_n>
```

&nbsp;

### <u>For Loops In R</u>

&nbsp;

For loops is a compact way to repeatedly run lines of code while changing variable values. The for-loop format in R would look something like:

&nbsp;

```
for (index_var in min:max){
    <run code here>
}
```

&nbsp;

A simple example of a for loop in printing numbers from 1 to 10.

```{r}
for (i in 1:10){
print(i)
}
```

&nbsp;

---

&nbsp;

### <u>The Fibonacci Function In R</u>

&nbsp;

The code for the Fibonacci function can be found below. The user puts in a single argument which should be a positive whole number (integer) that is at least 1 or greater. In the function, the if-else statement deals with cases for numbers less than 1 and for non-whole numbers (by making sure the fractional part of the number is 0).

Since the Fibonacci sequence relies on the previous two numbers in the sequence (with 1 and 1 as the first two numbers), a for loop is used here. The n-th Fibonacci number in the sequence is represented by fibonacci_vect[n] which is used in the paste0() string concatenation function.


```{r}
# Fibonacci Sequence in R

# 1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, and so on.

# A Function Which ouptuts the n-th Fibonacci number (n >= 1 & n is natural number):

fibonacci_funct <- function(n) {
 
 if (n < 1){
 print("Please choose a whole number that is 1 or greater")
 } else if (n == 1) {
 return(print("The first Fibonacci number is 1."))
 } else if (n == 2) {
 return(print("The second Fibonacci number is 2."))
 } else if ((n >= 3) && (n %% 1 == 0)){
 
 fibonacci_vect = c() #Initalize empty vector
 
 # Initalize first two numbers in Fibonacci sequence:
 
 fibonacci_vect[1] <- 1 
 
 fibonacci_vect[2] <- 1
 
 for (i in 3:n){
 fibonacci_vect[i] = fibonacci_vect[i - 2] + fibonacci_vect[i - 1]
 }
 
 # Create output message with paste0() concatenation:
 
 msg <- paste0("Position: ", n, ", Fibonacci Number: ", fibonacci_vect[n])
 
 return(msg) # Output n-th Fibonacci number as a message
 } else # Non-whole number case.
 print("Invalid number. Please enter in a whole number that is 1 or greater.")
}
```

&nbsp;

### <u>Test Cases</u>

&nbsp;

To make sure that the function works and is error free, some testing is required. I want to test cases where the numbers in the function call are negative, 1, 2, and numbers with a fractional part (non-whole numbers).


```{r}
# Test Cases:

fibonacci_funct(2)

fibonacci_funct(-1)

fibonacci_funct(1)

fibonacci_funct(0)

fibonacci_funct(2.4)

fibonacci_funct(3)

fibonacci_funct(6)

fibonacci_funct(10)
```

&nbsp;

### <u>References</u>

&nbsp;

https://stackoverflow.com/questions/3476782/check-if-the-number-is-integer

https://www.programiz.com/r-programming/if-else-statement