---
title: "rmath-randomwalks"
output: html_document
---

&nbsp;

Hi. This page is about using R to simulate random walks. Results are then plotted in R with the `ggplot2` package.

&nbsp;

### <u>Sections</u>

&nbsp;

* <a href="#rwalk">The Random Walk</a>

* <a href="#ggplot2">Plotting The Random Walk With ggplot2</a>

* <a href="#method">A More Efficient Method</a>

* <a href="#function">The Final Random Walk Plotting Function</a>

* <a href="#refs">References</a>

&nbsp;

<a name="rwalk"></a>

### <u>The Random Walk</u>

&nbsp;

In here, the random walk is referred as the symmetric random walk. A random walk is a stochastic (random) process where it is based on a sequence of random variables indexed by time.

Suppose we have a random variable $X_t$ at time $t$ where

&nbsp;

$$X_t = +1 \text{ with a probability of 0.5 }$$

&nbsp;

and

&nbsp;

$$X_t = -1 \text{ with a probability of 0.5 }$$

&nbsp;

The random walk is the the sum of these $X_t$ random variables and can be seen as the position of the random process at time $t$.

In math notation, this sum can be represented as:

&nbsp;

$$M_t = \displaystyle\sum_{t=1}^{n} X_t$$

&nbsp;

with $M_0 = 0$ (Starting point at 0.)

&nbsp;

<a name="ggplot2"></a>

### <u>Plotting The Random Walk With ggplot2</u>

&nbsp;

Once how the random walk works is understood, you can code a random walk simulation into R with the ggplot2 data visualization package.

Assuming ggplot2 is installed into R, loading ggplot2 can be done by typing in:

&nbsp;

```
library(ggplot2)
```

&nbsp;

**Creating The Random Walk Function**

&nbsp;

In R, I create a random walk function which takes in a time (in seconds) and outputs a vector of random walk positions at each time $t$. This output is:

&nbsp;

$$M_t = \displaystyle\sum_{t=1}^{n} X_t$$

&nbsp;

with $M_0 = 0$ (Starting point at 0.)

Since a sum is used, a for loop is used in the function.

To model the random variable which is +1 or -1 with a probability of 0.5 each, the `sample()` function is used.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# Random Walk Plot In R With ggplot2

# Resouce: https://stackoverflow.com/questions/21991130/simulating-a-random-walk

# The random walk is a running total/sum based on a random variable
# that is +1 or -1 depending on the outcome.

# In here, we deal with a symmetric random walk where the expected value of the running total/sum
# is zero and the random variable has an equal chance of being +1 or -1.

library(ggplot2)

# Random Walk Function:
# Input: Time in seconds; Output: The sequence of sums at each time point with M_0 = 0.

random_walk <- function(t = 1){
  if (t <= 0 | t %% 1 != 0){
    print("Please enter a whole number that is 1 or greater.")
  } else {
  
  # Initialize
  time_vect <- 0:t
  sum_vect <- c(0) 
  
  # Remember that M_0 = 0 meaning that the sum at t = 0 is 0
  for (i in 1:t){
    sum_vect[i + 1] = sum_vect[i] + sample(x = c(-1, 1), size = 1, prob = c(0.5, 0.5))
  }
  
  return(sum_vect)
}
}
```

&nbsp;

Once this function has been implemented into R, this function can be used to help in plotting a random walk path.

&nbsp;

```{r}
time = 5

rw5 <- random_walk(5)

data_rw5 <- data.frame(0:5, rw5)

data_rw5

colnames(data_rw5) <- c("t", "Position") #Rename columns

data_rw5
```

&nbsp;

Once the coding for the random walk function is done, the ggplot2 code portion is not that bad. Here is the code and plot.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# Ggplot with time on horizontal axis, and running total on vertical axis.

ggplot(data_rw5, aes(x = t, y = Position)) + 
  geom_point() +
  geom_line() +
  labs(x = "\n t", y = "Position \n", title = paste0("Random Walk With ", time, " Steps \n")) + 
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12)) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dotted")
```

&nbsp;

The main pieces in the above code lie in this portion:

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
ggplot(data_rw5, aes(x = t, y = Position)) + 
  geom_point() +
  geom_line()
```

&nbsp;

Code after geom_line is for labels, fonts and for the red horizontal dotted line at 0 for position. This red dotted line represents the mean line of the symmetric random walk.

&nbsp;

<a name="method"></a>

### <u>A More Efficient Method</u>

&nbsp;

Instead of having the for loop portion in the random walk function above, you can use the `cumsum()` function (short for cumulative sum).

The `cumsum()` function outputs a vector of sums which is similar to the sum/Sigma notation mentioned above.

This portion of code:

&nbsp;

```
# Remember that M_0 = 0 meaning that the sum at t = 0 is 0
  for (i in 1:t){
    sum_vect[i + 1] = sum_vect[i] + sample(x = c(-1, 1), size = 1, prob = c(0.5, 0.5))
  }
```

&nbsp;

can be replaced by this code (with output) below.

&nbsp;

```{r}
# Using the cumsum() function to generate random walk more efficently:

time <- 5

rw_psn <- c(0, cumsum(sample(x = c(-1, 1), size = time, prob = c(0.5, 0.5), replace = TRUE)))

rwalk_data5 <- data.frame(0:5, rw_psn)

rwalk_data5
```

&nbsp;

Here is the full code and output with the cumsum() function.

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# Using the cumsum() function to generate random walk more efficiently:
 
time <- 5

rw_psn <- c(0, cumsum(sample(x = c(-1, 1), size = time, prob = c(0.5, 0.5), replace = TRUE)))

rwalk_data5 <- data.frame(0:5, rw_psn)

rwalk_data5

colnames(rwalk_data5) <- c("t", "Position")

# Ggplot with time on horizontal axis, and running total on vertical axis.

ggplot(rwalk_data5, aes(x = t, y = Position)) + 
  geom_point() +
  geom_line() +
  labs(x = "\n t", y = "Position \n", title = paste0("Random Walk With ", time, " Steps \n")) +
  theme(plot.title = element_text(hjust = 0.5), 
  axis.title.x = element_text(face="bold", colour="blue", size = 12),
  axis.title.y = element_text(face="bold", colour="blue", size = 12)) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dotted")
```

&nbsp;

Notice that the random walk path is different from the one before as it is a different random walk path realization.

&nbsp;

<a name="function"></a>

### <u>The Final Random Walk Plotting Function</u>

&nbsp;

This final random walk plotting function combines items from the previous two sections. I have included the full code and one example.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
## Making It As A Random Walk Plotting Function:
# Input: Time in seconds; Output: The sequence of sums at each time point with M_0 = 0.


random_walk_plot <- function(t = 1, col_line = "black"){
 if (t <= 0 | t %% 1 != 0){
 print("Please enter a whole number that is 1 or greater.")
 } else {
 
 # Random Walk Steps:
 psn_vector <- c(0, cumsum(sample(x = c(-1, 1), size = t, prob = c(0.5, 0.5), replace = TRUE)))
 
 # Put results into a data frame:
 rwalk_df <- data.frame(0:t, psn_vector)
 
 colnames(rwalk_df) <- c("t", "Position") #Rename columns
 
 # Ggplot with time on horizontal axis, and running total on vertical axis.
 
 ggplot(rwalk_df, aes(x = t, y = Position)) + 
 geom_point() +
 geom_line(colour = col_line) +
 labs(x = "\n t", y = "Position \n", title = paste0("Random Walk With ", t, " Steps \n")) + 
 theme(plot.title = element_text(hjust = 0.5), 
 axis.title.x = element_text(face="bold", colour="blue", size = 12),
 axis.title.y = element_text(face="bold", colour="blue", size = 12)) +
 geom_hline(yintercept = 0, colour = "red", linetype = "dotted")
 
 }
}

# Function calls:

random_walk_plot(t = 20, col_line = "green")
```


&nbsp;

<a name="refs"></a>

### <u>References</u>

&nbsp;

* R Graphics Cookbook By Winston Chang
* https://stackoverflow.com/questions/21991130/simulating-a-random-walk