---
title: "Simulating White Noise"
output: html_document
---

&nbsp;

Hi. This page focuses on simulating white noise in R. White Noise is a very simple stationary process in the field of time series analysis. Weak white noise has a fixed constant mean over time and a fixed constant variance and there is no correlation over time.

I use Datacamp Slides as my reference but the plots are in base R graphics. The plots here are generated with `ggplot2`.

To simulate observations from a White Noise process, the `arima.sim()` function is used. Make sure to use `model = list(order = c(0, 0, 0)`.

&nbsp;

```{r}
# Simulating White Noise (Time Series)

# Reference: Datacamp Slides

# In the slides, a base R graphics plot is used. I want to use ggplot2 as ggplot2 looks
# nicer.

library(ggplot2)

# White Noise is a very simple stationary process.

# Weak white noise has a fixed constant mean, a fixed constant variance and
# no correlation over time.

#### Simulate 100 observations in White Noise model.

wn_sim_100 <- arima.sim(model = list(order = c(0, 0, 0)), n = 100)

head(wn_sim_100)

ts_wn100 <- data.frame(Time = seq(0, 99, 1), WN = wn_sim_100)

head(ts_wn100); tail(ts_wn100)
```

&nbsp;

From the `ts_wn100` data frame, a plot can be generated with the help of `ggplot2` graphics.

&nbsp;

```{r, echo = TRUE, fig.width=5, fig.height=5} 
# Ggplot2 line plot of simulated white noise observations:

ggplot(data = ts_wn100, aes(x = Time, y = WN)) + 
  geom_line() +
  labs(x = "\n Time ", y = "WN \n", title = "Simulated Observations From A White Noise Process \n") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(face="bold", colour="darkblue", size = 12),
        axis.title.y = element_text(face="bold", colour="darkblue", size = 12))
```

&nbsp;


**Adding More Observations In White Noise Simulation**

&nbsp;

```{r, echo = TRUE, fig.width=5.5, fig.height=5.5} 
### Simulate 10000 observations in White Noise model.

wn_sim10000 <- arima.sim(model = list(order = c(0, 0, 0)), n = 10000)

head(wn_sim10000)

ts_wn10000 <- data.frame(Time = seq(0, 9999, 1), WN = wn_sim10000)

head(ts_wn10000); tail(ts_wn10000)

# Ggplot2 line plot of simulated white noise observations:

ggplot(data = ts_wn10000, aes(x = Time, y = WN)) + 
  geom_line() +
  labs(x = "\n Time ", y = "WN \n", title = "Simulated Observations From A White Noise Process \n") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(face="bold", colour="darkblue", size = 12),
        axis.title.y = element_text(face="bold", colour="darkblue", size = 12))
```

&nbsp;

## A White Noise Process With Drift And A Different Variance

&nbsp;

The simulations above dealt with a White Noise process with a mean of 0 and a variance of 1. This next part deals with a white noise model with a mean of 2 and a variance of 4 (or standard deviation of 2).

&nbsp;

```{r, echo = TRUE, fig.width=4.5, fig.height=4.5} 
# Simulating observations from white noise model with different means and variance.

# Simulate 50 observations from the WN model with mean = 2, sd = 2 (var = 4)

wn_2 <- arima.sim(model = list(order = c(0, 0, 0)), n = 50, mean = 4, sd = 2)

head(wn_2)

ts_wn_2 <- data.frame(Time = seq(0, 49, 1), WN = wn_2)

head(ts_wn_2); tail(ts_wn_2)

# Ggplot2 line plot of simulated white noise observations:

ggplot(data = ts_wn_2, aes(x = Time, y = WN)) + 
  geom_line() +
  labs(x = "\n Time ", y = "WN \n", title = "Simulated Observations From A \n White Noise Process (Mean = 4, SD = 2) \n") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(face="bold", colour="darkblue", size = 12),
        axis.title.y = element_text(face="bold", colour="darkblue", size = 12))

# Fit second white noise model with arima():

arima(wn_2, order = c(0, 0, 0))

# Mean and variance of white noise:

mean(wn_2)

var(wn_2)
```

