---
title: "Working With A Two By Two Table In R"
output: html_document
---

&nbsp;

Hello everyone. This page is about working with a two by two table in the statistical programming language R.

&nbsp;

### <u>Sections</u>

&nbsp;

* <a href="#data">Creating Sample Data</a>
* <a href="#table">A Two By Two Contingency Table</a>
* <a href="#poisson">Poisson Regression</a>
* <a href="#refs">References</a>

&nbsp;

<a name="data"></a>

### <u>Creating Sample Data</u>

&nbsp;

I start with creating sample (fake) data where males and females are surveyed whether or not they like sushi or not.

&nbsp;

```{r}
# Contingency Tables In R
# Book: Extending The Linear Model With R By Julian J Faraway

# Creating a Sample Table: Do You Like Sushi By Gender?
# gl() generates factor levels

library(ggplot2)

counts <- c(19, 24, 18, 21)
gender <- gl(n = 2, k = 1, length = 4, labels = c("Male", "Female"))
interest <- gl(n = 2, k = 2, length = 4, labels = c("Yes", "No"))

survey_data <- data.frame(counts, gender, interest)

survey_data
```

&nbsp;

From the survey data, you can easily create bar graphs with the ggplot2 package in R.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4}
#### ggplot2 Graphs


# Data Visualization Of Contingency Table With ggplot2 (Stacked Bar Graph):

ggplot(data = survey_data, aes(x = interest, y = counts, fill = gender)) + 
  geom_bar(stat = "identity") + 
  labs(x = "\n Answer", y = "Counts \n", 
       title = "Interest In Sushi By Gender \n",
       fill = "Gender") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12),
        legend.position = "bottom")
```

&nbsp;

The above plot is a stacked bar graph. An alternative to the above would be side by side bar graphs.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4}
# Data Visualization Of Contingency Table With ggplot2 (Side By Side Bar Graph):

ggplot(data = survey_data, aes(x = interest, y = counts, fill = gender)) + 
  geom_bar(stat = "identity", position = "dodge", colour = "black") + 
  scale_fill_brewer(palette = "Pastel1") +
  labs(x = "\n Answer", y = "Counts \n", 
       title = "Interest In Sushi By Gender \n",
       fill = "Gender") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(face="bold", colour="blue", size = 12),
        axis.title.y = element_text(face="bold", colour="blue", size = 12),
        legend.position = "bottom")
```

&nbsp;

I have also changed the colour palette to mix things up.

&nbsp;

<a name="table"></a>

### <u>A Two By Two Contingency Table</u>

&nbsp;

Instead of the long format from the beginning, you can display the table as a two by two contingency table.

&nbsp;

```{r}
### Contingency Tables (2 by 2 Case)

conting_table <- xtabs(counts ~ gender + interest)

conting_table
```

&nbsp;

From this contingency table, you can create a mosaic plot.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4}
# Mosaic Plot (Base R):

mosaicplot(conting_table, color = c("red", "green"), main = "Mosaic Plot",
           xlab = "Gender", ylab = "Answer")
```

&nbsp;

Since the counts are really close to each other, it is hard to see a difference between the tile sizes.

An alternate moasic plot comes from the vcd package in R.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# Mosaic Plot (vcd package):

library(vcd)

mosaic( ~ gender + interest , data = conting_table,
        highlighting = "gender", highlighting_fill=c("lightblue", "pink"))
```

&nbsp;

Other than the colours and labels, this mosaic plot does not look that much different. Also, I have not figured out how to adjust the label titles and such.

&nbsp;

<a name="poisson"></a>

### <u>Poisson Regression</u>

&nbsp;

The counts are at least zero (non-negative) and are whole numbers. When dealing with a two by two table, linear regression does not really work. With this data, a Poisson regression model is used.

In R, the `glm()` function is used where glm stands for generalized linear model. Make sure to indicate `family = "poisson"` in the `glm()` function.

&nbsp;

```{r}
# Poisson Regression 

poisson_model <- glm(counts ~ gender + interest, family = "poisson", data = survey_data)

summary(poisson_model)
```

&nbsp;

<a name="refs"></a>

### <u>References</u>

&nbsp;

* R Graphics Cookbook By Winston Chang
* Extending The Linear Model By Julian J Faraway
