---
title: "A Sideways Bar Graph Example In R"
output: html_document
---

&nbsp;

Hi. I have been playing around R's ggplot2 data visualization package and I have learned how to do sideways bar graphs.

&nbsp;

### <u>Table Of Contents</u>

&nbsp;

* References
* Favourite Colours Survey Data
* From Regular Bar Graph To A Sideways Bar Graph
* Retaining Order In Factors In A Bar Graph
* Sorted Sideways Bar Graph
* Adding Labels To The Sorted Sideways Bar Graph

&nbsp;

### <u>References</u>

&nbsp;

* http://rstudio-pubs-static.s3.amazonaws.com/7433_4537ea5073dc4162950abb715f513469.html
* R Graphics Cookbook by Winston Chang (2012)

&nbsp;

### <u>Favourite Colours Survey Data</u>

&nbsp;

I will illustrate how to create a sideways bar graph using made up survey data. This data will be based on people's favourite colours.

&nbsp;

```{r}
# Load packages:

library(ggplot2)

# Favourite Colour Survey (Made-Up/Fake Data):

colour_choices <- c("Blue", "Red", "Green", "White", "Black", "Orange", 
                    "Purple", "Pink", "Yellow", "Brown")

colour_counts <- c(35, 26, 33, 19, 20, 15, 12, 24, 30, 29)

colour_table <- data.frame(colour_choices, colour_counts)

# Preview table:

colour_table
```

&nbsp;


The next lines of code consist of check the structure of the data, renaming the column names and computing the total number of people in the survey.

```{r}
# Check data structure:

str(colour_table)

# Rename columns:

colnames(colour_table) <- c("Colour", "Count")

# Total Number in Survey:

(total_num <- sum(colour_table[, 2]))
```

&nbsp;

### <u>From Regular Bar Graph To A Sideways Bar Graph</u>

&nbsp;

The code for a regular (vertical) bar graph in R using ggplot2 would look like this.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
## ggplot bar Graph:

ggplot(data = colour_table, aes(x = Colour, y = Count)) +
 geom_bar(stat = "identity", width = 0.75) +
 labs(x = "\n Colour Choice", y = "Count \n", title = "Favourite Colours Survey Results \n") +
 theme(plot.title = element_text(hjust = 0.5), 
 axis.title.x = element_text(face="bold", colour="red", size = 12),
 axis.title.y = element_text(face="bold", colour="red", size = 12))
```

&nbsp;

To convert the above bar graph into a sideways one, just simply add coord_flip() after ggplot().

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
## Sideways Bar Graph (Add coord_flip())

ggplot(data = colour_table, aes(x = Colour, y = Count)) +
 geom_bar(stat = "identity", width = 0.75) +
 coord_flip() +
 labs(x = "\n Colour Choice", y = "Count \n", title = "Favourite Colours Survey Results \n") +
 theme(plot.title = element_text(hjust = 0.5), 
 axis.title.x = element_text(face="bold", colour="red", size = 12),
 axis.title.y = element_text(face="bold", colour="red", size = 12))
```

&nbsp;

Notice that from bottom to top the colours are in ABC order. This is different than what we had earlier with what we had with `colour_choices()`.

&nbsp;

```{r}
colour_choices <- c("Blue", "Red", "Green", "White", "Black", "Orange", 
"Purple", "Pink", "Yellow", "Brown")
```

&nbsp;

### <u>Retaining Order In Factors In A Bar Graph</u>

&nbsp;

We can retain the colour choices order we want with a minor fix. This fix can be done on the first column by convert the first column as a factor but with setting the vector colour_choices in the levels argument of `factor()`.

&nbsp;

```{r}
## Colours are in ABC order, not in order as defined in the beginning. Some fixes needed.

# Fix first column:

colour_table$Colour <- factor(colour_table$Colour, levels = colour_choices)
```

&nbsp;

The resulting sideways bar graph code is similar to the previous one. The resulting sideways bar graph will be different on the vertical axes with the colour choices.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
ggplot(data = colour_table, aes(x = Colour, y = Count)) +
 geom_bar(stat = "identity", width = 0.75) +
 coord_flip() +
 labs(x = "\n Colour Choice", y = "Count \n", title = "Favourite Colours Survey Results \n") +
 theme(plot.title = element_text(hjust = 0.5), 
 axis.title.x = element_text(face="bold", colour="red", size = 12),
 axis.title.y = element_text(face="bold", colour="red", size = 12))
```

&nbsp;


### <u>Sorted Sideways Bar Graph</u>

&nbsp;

If we want to sort our bars from largest to smallest, we need to reorganize the factors in a specific way.

&nbsp;

```{r}
## Sorting the colours from most popular to least popular sideways bar graph:

# We fix it by using the factor function again but with a modification.
# Reference: http://rstudio-pubs-static.s3.amazonaws.com/7433_4537ea5073dc4162950abb715f513469.html
# order(colour_table$Count) puts the row numbers from smallest to largest:

colour_table$Colour <- factor(colour_table$Colour, 
                              levels = colour_table$Colour[order(colour_table$Count)])
```

&nbsp;

The `order(colour_table$Count)` part outputs the positions of the Count column in colour_table from largest to smallest. Then `colour_table$Colour[order(colour_table$Count)]` outputs the colours with the largest counts first to the less frequent colour.

&nbsp;

Here is the code for the sorted sideways bar graph with the output.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
ggplot(data = colour_table, aes(x = Colour, y = Count)) +
 geom_bar(stat = "identity", width = 0.75) +
 coord_flip() +
 labs(x = "\n Colour Choice", y = "Count \n", title = "Favourite Colours Survey Results \n") +
 theme(plot.title = element_text(hjust = 0.5), 
 axis.title.x = element_text(face="bold", colour="red", size = 12),
 axis.title.y = element_text(face="bold", colour="red", size = 12))
```

&nbsp;

### <u>Adding Labels To The Sorted Sideways Bar Graph</u>

&nbsp;

To make the bar graphs a bit more informative, labels can be added. The `geom_text()` function after `ggplot()` enables labels.

```{r, echo = TRUE, fig.width=4, fig.height=4}
## Adding labels (Add geom_text()):

ggplot(data = colour_table, aes(x = Colour, y = Count)) +
 geom_bar(stat = "identity", width = 0.75) +
 coord_flip() +
 geom_text(aes(label = Count), hjust = 1.2, colour = "white", fontface = "bold") +
 labs(x = "\n Colour Choice \n", y = "\n Count \n", title = "\n Favourite Colours Survey Results \n") +
 theme(plot.title = element_text(hjust = 0.5, size = 15), 
 axis.title.x = element_text(face="bold", colour="red", size = 12),
 axis.title.y = element_text(face="bold", colour="red", size = 12))
```

