---
title: "Side By Side Bar Graphs In R & ggplot2"
output: html_document
---

&nbsp;

### <u>Table Of Contents</u>

&nbsp;

* Simulated Coin Flip Data
* Side By Side Bar Graphs
* Using facet_grid()
* References

&nbsp;


### <u>Simulated Coin Flip Data</u>

&nbsp;

The `ggplot2` package is first loaded into R.

&nbsp;

```{r}
library("ggplot2")
```

&nbsp;

We have two players A and B who each an unfair coin. This unfair coin has a 30% chance of getting heads and a 70% chance of getting tails. Both players flip this coin 1000 times each. This can be simulated in R using the `sample()` function.

&nbsp;

```{r}
# Flip a unfair coin 1000 times 

flip_results_A <- sample(c("Heads", "Tails"), size = 1000, replace = TRUE, prob = c(0.3, 0.7))
flip_results_B <- sample(c("Heads", "Tails"), size = 1000, replace = TRUE, prob = c(0.3, 0.7))
```

&nbsp;

We can see the results of the coin flips using the `table()` function.

&nbsp;

```{r}
# Get counts of the die rolls:

table(flip_results_A) ; table(flip_results_B)
```

&nbsp;

The next lines of code is about putting the results together in a data table format.

&nbsp;

```{r}
# Merge the results from Player A and Player B together:
# Format: Player A | Player A flip results (1000 rows)
# Player B | Player B flip results (1000 rows)

column1 <- c(rep("Player A", 1000), rep("Player B", 1000))
column2 <- c(flip_results_A, flip_results_B)

outcome_data <- cbind(column1, column2)

flip_results <- data.frame(Player = factor(column1, levels = c("Player A", "Player B")),
                Result = column2)

# A peek of the merged data:
head(flip_results); tail(flip_results)
 
```

&nbsp;

These coin flip results can be converted into a two by two table as a different visual. For graphing purposes, this 2 by 2 table is converted into a data frame.

&nbsp;

```{r}
table(flip_results)

# The two by two table into a data frame for plotting:

results <- data.frame(table(flip_results))
```

&nbsp;

### <u>Side By Side Bar Graphs</u>

&nbsp;

To obtain side by side bar graphs in `ggplot2`, we need a lot of parts on top of the `ggplot()` command.

* geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75) gives the side by side bar graphs

* ylim(0, 800) gives limits on the y-axis values

* The geom_text() line adds labels to the bar graphs. Note that position_dodge is needed as we used position dodge was used in geom_bar().

* labs() gives labels depending on what is specified

* The theme() function allows for additional aesthetic options such as a centered title and font sizes.

&nbsp;


```{r, echo = TRUE, fig.width=4, fig.height=4} 
# Plotting side by side bar graphs:

# http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/
# R Graphics Cookbook by Winston Chang Reference
# Result of heads or tails in x = axis, Counts as y axis, diff colours for each player.
# Put labels:

ggplot(data = results, aes(x = Result, y = Freq, fill = Player)) +
  geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75)  +
  ylim(0,800) +
  geom_text(aes(label = Freq), fontface = "bold", vjust = 1.5,
             position = position_dodge(.9), size = 4) +
  labs(x = "\n Coin Flip Outcome", y = "Frequency\n", title = "\n Coin Flip Results \n") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="red", size = 12),
        axis.title.y = element_text(face="bold", colour="red", size = 12),
        legend.title = element_text(face="bold", size = 10))
```

&nbsp;

### <u>Using facet_grid()</u>

&nbsp;


In the above plot, we have four bars in one graph where the Player A bars are beside the Player B bars. Adding facet_grid(. ~Player) will turn this bar graph plot in a way where there is one bar graph plot for Player A and another bar graph plot for player B right beside it.

&nbsp;

```{r, echo = TRUE, fig.width=4, fig.height=4} 
# Facet_grid:

ggplot(data = results, aes(x = Result, y = Freq)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  facet_grid(. ~Player)  +
  ylim(0,800) +
  geom_text(aes(label = Freq), fontface = "bold", vjust = 1.5, colour = "white", size = 4) +
  labs(x = "\n Coin Flip Outcome", y = "Frequency\n", title = "\n Coin Flip Results \n") +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.title.x = element_text(face="bold", colour="darkgreen", size = 12),
        axis.title.y = element_text(face="bold", colour="darkgreen", size = 12),
        legend.title = element_text(face="bold", size = 10),
        strip.background = element_rect(fill="lightblue", colour="black", size=1),
        strip.text = element_text(face="bold", size=rel(1.2)))
```

&nbsp;

Note that the `position_dodge()` options were removed as they are not needed. To highlight the Player A and Player B font text and the bars, the strip.background and strip.text options are used.

&nbsp;


### <u>References</u>

&nbsp;

* http://www.cookbook-r.com/Graphs/Bar_and_line_graphs_(ggplot2)/
* R Graphics Cookbook by Winston Chang (2012)